\titlea\titleb\titlec\titled\titlee\titlef\titleg

### 2016-09-01 oopsla-16/evaluation.tex
oopsla: clean up/remove outdated comments in main tex

	\item Does \quicksand~improve upon state-of-the-art MC?
		\begin{enumerate}
			\revision{\item} Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
			\revision{\item} Does \quicksand~find bugs faster
				%than SSS-MC
				in subset state spaces, even without data-race PPs?
			% Probably not... ICB is state of the art here.
			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
			\revision{
			\item Does \quicksand~provide more full verifications of bug-free programs more quickly?
			}
			\revisionx{
			\item How does the choice between HB and Limited HB affect bug-finding and verification performance?
			}
		\end{enumerate}
	\item Does MC improve the accuracy of data-race detection?
		\begin{enumerate}
		\item Does \quicksand~avoid false positives compared to a single-execution \revisionx{Limited HB} data race analysis?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% But to be fair, you have to count how many DRs are reported as "couldn't test these, check yourself" at the end.
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
				% to do (done) If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
				% Answer: Just 1. (for p2s at least; as of time of writing, pintos dr-falsenegs not run yet)
			\item Does \quicksand~find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
		\end{enumerate}
\end{enumerate}

@@

\subsection{Test Suite}
\label{sec:testsuite}
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from Carnegie Mellon's 15-410 operating systems class,
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's
\revision{CMSC 23000}~operating systems classes.
%
The P2 project comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and reader/writer locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
comprises priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided mutex, context-switch, and virtual memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
%Though not ``real world'' programs,
Both projects are quite complex:
the P2s average 1807 lines of code, %C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/daniel-deniz*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718 lines, for a total of 198,772 lines tested for this paper.

@@

We tested P2s with 6 multithreaded programs:
% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
% each tailored to exercise a different part of the P2 project
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for condition variables,
\paradise~for semaphores,
% XXX f u latex
and \rwldgr\\for r/w locks.
For \mxtest, \paradise, and \paraguay, we used the {\tt without\_function} command to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
We tested Pintoses with 3 programs from the class's provided test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented,
	so each test could only be run on a subset of the 78 submissions; see ``Number tested'' in Table~\ref{tab:drbugs}.
}.
The source code of all 9 test cases is available at
\url{https://github.com/bblum/oopsla-dataset}.
For all tests,
%we used \revisionz{preemption sealing (\sect{\ref{sec:landslide}})}
%{\tt without\_function}
%to blacklist PPs on {\tt malloc}'s internal lock.
\revisionz{we blacklisted PPs on {\tt malloc}'s internal lock
using \landslide's {\tt without\_function} command (\sect{\ref{sec:landslide}}).}
In total, our evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
% at least 157 of which could expose bugs.
at least \revisionz{181} of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt

%\begin{table}[t]
%	\begin{tabular}{l|l|l}
%			& QS bugs & SSS-MC bugs \\
%		\hline
%		\mxtest & eg 1000 & eg 0 \\
%		\bct & & \\
%		etc... & & \\
%		\hline
%		Total & & \\
%	\end{tabular}
%	\caption{Comparison of all bugs found, broken down by test case, among all P2s (top 6) and Pintoses (bottom 2)}
%	\label{tab:allbugs}
%\end{table}
%
%\begin{table}[t]
%	\small
%	\begin{tabular}{l|l|l||l|l}
%	& QS bug & \begin{tabular}{c} SSS-MC \\ completed\end{tabular}
%	& QS bug & \begin{tabular}{c}SSS-MC \\ timeout \end{tabular} \\
%		\hline
%		\mxtest & e.g. 5 & 10 & 0 & 0 \\
%		\bct & & & & \\
%		etc... & & & & \\
%		\hline
%		Total & & & & \\
%	\end{tabular}
%	\caption{Bugs requiring data-race PPs to expose, found by \quicksand~but missed by the single-state-space approach.}
%	\label{tab:drbugs}
%\end{table}

@@

\begin{table*}[t]
	\begin{center}
		\small
		\revisionx{
	\begin{tabular}{r|c||c|c|c|c|c||c|c|c|c||c|c}
		\multicolumn{2}{c||}{} & \multicolumn{5}{c||}{\bf \revz{Total bugs}} & \multicolumn{4}{c||}{\bf \revz{Data-race bugs}} \\
		% ---
		& {\bf Num.} & \multicolumn{2}{c|}{\bf \revz{\quicksand}} & \multicolumn{3}{c||}{\bf \revz{Control (SSS-MC)}}
		& \multicolumn{2}{c|}{\bf \revz{Limited HB}} & \multicolumn{2}{c||}{\bf \revz{Pure HB}} &
		%& {\bf LHB DR} & {\bf Nondet.} & & &
		{\bf Mutual} & {\bf Avg. tested} \\
		% ---
		{\bf Test} & {\bf tested} & {\bf LHB} & {\bf PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem}
		& {\bf All} & {\bf Nondet.} & {\bf \revz{All}} & {\bf \revz{Nondet.}}
		& {\bf timeouts} & {\bf subset SSes} \\
		% ---
		\hline
				% #test  qs-lhb  qs-phb           icb             dpor    everyw  dronly  nondets(LHB)    dronly(PHB)    nondets(LHB) mutualtimeout    comp.SSes
		{\tt bcast}	& 79	& 8	& 8		& 5		& 6	& 7	& 2	& \revz{1}	& \revz{2}	& \revz{1}	& 7		& 112.3	\\
		{\tt join} 	& 79	& 23	& 20		& 13		& 13 	& 14	& 11	& \revz{4}	& \revz{7}	& \revz{3}	& \revz{12}	& \revz{69.7}	\\
		{\tt mx} 	& 79	& 10	& 9		& 1		& 1	& 12	& 9	& \revz{1}	& \revz{8}	& \revz{1}	& 0		& -	\\
		{\tt sem} 	& 79	& 17	& 16		& 12		& 11	& 12	& 7	& \revz{3}	& \revz{6}	& \revz{2}	& \revz{50}	& \revz{77.4}	\\
		{\tt signal} 	& 79	& 10	& 8		& 5		& 5	& 11	& 6	& \revz{1}	& \revz{3}	& \revz{2}	& 45		& \revz{59.6}	\\
		{\tt rwlock} 	& 79	& 27	& 26		& 25		& 23	& 28	& 4	& \revz{1}	& \revz{3}	& \revz{0}	& \revz{44}	& \revz{86.3}	\\
		\hline
		{\tt sched} 	& 59	& 7	& 7		& 1		& 1	& 8	& 6	& \revz{4}	& \revz{6}	& \revz{6}	& 2		& 13.0	\\
		{\tt alarm} 	& 44	& 21	& 12		& 16		& 5	& 29	& 17	& \revz{1}	& \revz{7}	& \revz{6}	& \revz{17}	& \revz{7.8}	\\
		{\tt wait} 	& 52	& 30	& \revz{26}	& \revz{24}	& 23	& 1	& 7	& \revz{2}	& \revz{2}	& \revz{0}	& \revz{15}	& \revz{33.8}	\\
		\hline
		{\bf Total}	& 629	& 153	& \revz{132}	& \revz{102}	& 88	& 122	& 69	& \revz{15}	& \revz{44}	& \revz{21}	& \revz{192}	& \revz{65.8} 	\\
	\end{tabular}
		}
	\end{center}
	\caption{Summary of bugs
	%and data races
	found by each test program.
		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
		``\revisionz{Data-race} bugs'' counts among \revisionz{\quicksand's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
	among those, \revisionz{the ``Nondeterministic'' columns show} how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
		}
	\label{tab:drbugs}
\end{table*}

{\bf Finding new data-race bugs.}
%We quickly pull ahead of SSS-MC, and ultimately conclude with 179\% as many bugs in total.
%The break-even point is at a negligible 90 seconds.
\revision{Compared to \revisionx{SSS-MC-ICB (the fastest among the control experiments),} \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds.
\revisionx{Compared to SSS-MC-Shared-Mem (the best SSS-MC approach in the long term), \quicksand's Limited HB version} ultimately concludes 10 CPU-hours with \revisionx{125\%} as many bugs in total.
\revision{Before the break-even point at 200 seconds,
\quicksand~lags behind \revisionx{SSS-MC-ICB} due to additional start-up overhead from its tenfold parallelism.
However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
Figure~\ref{fig:dowefindbugsfaster}(b) gives \quicksand~full credit for its inherent parallelism:
with a 10 core allocation, it outperforms SSS-MC
for any fixed budget of wall-clock time.}
\revisionx{After 1 hour of wall-clock time, tenfold \quicksand~performs 158\% as well as SSS-MC-ICB.}

@@

The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test program.
In \mxtest, in which we do not trust the lock implementation's correctness,
\revisionx{SSS-MC-ICB and SSS-MC-DPOR}
found dramatically fewer bugs (just 1)\footnote{
	The one bug SSS-MC found was in a fully-assembler lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
%Though for many applications of MC, assuming a correct lock implementation is sufficient,
Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.

%% RIP this.
%Furthermore, we plotted another line from this dataset, QS-no-DR-bugs,
%which represents only the bugs found in state spaces without any data-race PPs (like QS-no-DR-PPs, but paying any overhead for using data-race PPs).
%Intuitively, this line shows that for programs with only benign data races,
%\quicksand~can afford the extra overhead of verifying them while still slightly edging out SSS-MC.

% Future work (for thesis maybe?) Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
{\bf Finding the same bugs faster.}
%To test whether Iterative Deepening is effective even for MC domains without data races, %such as message-passing distributed systems,
\revision{The QS-Sync-Only experiment tests whether}~Iterative Deepening is effective even for MC domains without data races.
\revision{When \quicksand~ignores all data-race candidates,
its results are competitive with SSS-MC-DPOR, but SSS-MC-ICB outperforms it.
This is unsurprising: the seed subsets of PPs
QS-Sync-Only is limited to
(\sect{\ref{sec:initial-pp}})
are much less flexible than ICB's preemption strategy (\sect{\ref{sec:overview-sssmc}}).
This result suggests that in future work, \quicksand~should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.}
%\footnote{
%Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
%we ran these tests for 2.5 hours on 4 CPUs each.
%Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.

%Even though SSS-MC mostly catches up to it by the end of the 10-hour budget,
%and is faster in the first 60 seconds due to less start-up overhead,
%QS-no-DR-PPs finds more of the bugs sooner thereafter.
%Hence, for modest CPU budgets,
%\quicksand~is likely to find bugs SSS-MC will miss.
%and for more ambitiously-sized tests,
%programmers can be more confident in the verification provided when \quicksand~times out with no bug found.
%We conclude that for smaller arbitrary CPU budgets, especially less than 1 hour,
%Iterative Deepening is likely to find bugs SSS-MC will miss.
%Moreover, it is also easy to imagine scaling up the size of each test case to test ,
%using more threads or longer sequences of API calls.
%We hope that is compelling even to users willing to spend many CPU-hours on testing.
%These results show explicitly that for arbitrary CPU-time budgets
%\footnote{The initial perfect overlap between QS-*-HB and QS-Sync-Only indicates how long it takes before the first data-race bug is found.}
%even after the extra overhead of verifying them, \quicksand~still slightly edges out SSS-MC

@@

{\bf Partial verification.}
When a MC job times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
%Beyond using the state space estimator to guess the percent coverage,
%We know of no approach to quantify the probability
%that a bug would be exposed by an untested interleaving,
While recent work \cite{randomized-scheduler} attempts to quantify the probability
\revisionz{that a bug remains in some untested interleaving},
%that an untested interleaving would expose a bug,
\quicksand~instead reports which subsets of PPs resulted in state spaces that did complete in time.
%\quicksand~does not attempt any such quantitative guarantee,
%but can at least report
On \revisionz{229} tests, \revisionx{SSS-MC-ICB} timed out after 10 hours with no bugs found.
Among these tests, \quicksand~found bugs in \revisionz{37}.
For the other \revisionz{192}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
%and the number of unique PPs among them.
%In 6 cases, \quicksand~also failed to complete anything; beyond these,
%between 1 and 253 state spaces were completed for each test
%Between 0 and 233 unique PPs were tested among some completed subsets, with a mean of 27.2 and median of 7.5.
These completions guarantee that, if the test program could expose a bug,
% Justify implicit hypothesis: Sync PPs + DR PPs = all possible relevant PPs.
it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
\revision{
%While obviously not as strong as full verification,
Prior work \cite{sealing} has argued the value of similar {\em compositional testing} when full verification is intractable,
deferring to the user's expertise to judge the value of each subset of PPs verified.
}

%Hence, in Table~\ref{tab:drbugs} we count how often \quicksand~uncovered a bug only in state spaces which included data-race PPs, while
%
%In Table~\ref{tab:allbugs} ....

@@

%Though SSS-MC-Shared-Mem appears to outperform \quicksand, its safety guarantee is unsound,
%as many tests break its assumption of stack-allocated memory being private.
%For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
%While SSS-MC-Shared-Mem must avoid excessive PPs on every private stack access for performance,
%\quicksand~detects these dynamically, without needing to discriminate between stack and heap.
%In total, SSS-MC-Shared-Mem claimed to verify 21 tests
%(across 7 of the 9 test cases)
%in which QS-DR ultimately found bugs,
%which further calls into question the rest of its completions.

%\quicksand's total verifications
%%before a given elapsed CPU consumption.
%\revision{against those provided by SSS-MC-Shared-Mem\footnote{
%	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
%	For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
%	SSS-MC-Shared-Mem's verifications are only sound in the absence of such sharing!
%	\quicksand~detects these dynamically, without needing to discriminate between stack and heap.}
%	%Expanding the statically-coded PPs to include stack accesses would make SSS-MC-Shared-Mem's performance even worse.
%	% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization. Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
%}.
%\quicksand~outperforms SSS-MC, being able to verify safety for
%{\bf 9999x}
%as many tests in 10 CPU-hours.
%We attribute this again to the overhead of the sheer number of statically-coded PPs that SSS-MC-Shared-Mem must use,
%as well as to the work which ICB necessarily repeats when increasing its preemption bound.}

@@

% Actually, while SSS-MC-Shared-Mem proved competitive with \quicksand~at finding data-race-related bugs,
% its heuristic to ignore threads' own stack accesses introduces unsoundness.
% <Move contents of footnote here.>
% Indeed, SSS-MC-Shared-Mem claimed a full verification on 999999 tests for which \quicksand~found bugs!
% % Hence, no SSS-MC strategy competes with quicksand for truly sound verification
% Hence, only SSS-MC-DPOR and SSS-MC-ICB compete with \quicksand for sound verification, and they are thoroughly outperformed.
% Between data-race bugs and verifications, \quicksand~is shown to provide the best of both worlds.

>>>>

	\item Does \quicksand~improve upon state-of-the-art MC?
		\begin{enumerate}
			\revision{\item} Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
			\revision{\item} Does \quicksand~find bugs faster
				%than SSS-MC
				in subset state spaces, even without data-race PPs?
			% Probably not... ICB is state of the art here.
			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
			\revision{
			\item Does \quicksand~provide more full verifications of bug-free programs more quickly?
			}
			\revisionx{
			\item How does the choice between HB and Limited HB affect bug-finding and verification performance?
			}
		\end{enumerate}
	\item Does MC improve the accuracy of data-race detection?
		\begin{enumerate}
		\item Does \quicksand~avoid false positives compared to a single-execution \revisionx{Limited HB} data race analysis?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% But to be fair, you have to count how many DRs are reported as "couldn't test these, check yourself" at the end.
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
			\item Does \quicksand~find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
		\end{enumerate}
\end{enumerate}

@@

\subsection{Test Suite}
\label{sec:testsuite}
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from Carnegie Mellon's 15-410 operating systems class,
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and University of Chicago's
\revision{CMSC 23000}~operating systems classes.
%
The P2 project comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and reader/writer locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
comprises priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided mutex, context-switch, and virtual memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
%Though not ``real world'' programs,
Both projects are quite complex:
the P2s average 1807 lines of code, %C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718 lines, for a total of 198,772 lines tested for this paper.

@@

We tested P2s with 6 multithreaded programs:
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for condition variables,
\paradise~for semaphores,
% XXX f u latex
and \rwldgr\\for r/w locks.
For \mxtest, \paradise, and \paraguay, we used the {\tt without\_function} command to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
We tested Pintoses with 3 programs from the class's provided test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented,
	so each test could only be run on a subset of the 78 submissions; see ``Number tested'' in Table~\ref{tab:drbugs}.
}.
The source code of all 9 test cases is available at
\url{https://github.com/bblum/oopsla-dataset}.
For all tests,
%we used \revisionz{preemption sealing (\sect{\ref{sec:landslide}})}
%{\tt without\_function}
%to blacklist PPs on {\tt malloc}'s internal lock.
\revisionz{we blacklisted PPs on {\tt malloc}'s internal lock
using \landslide's {\tt without\_function} command (\sect{\ref{sec:landslide}}).}
In total, our evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
at least \revisionz{181} of which will expose bugs under one or more MC trials. % QS bugs + bugs where SSSMC found but QS didnt

@@

\begin{table*}[t]
	\begin{center}
		\small
		\revisionx{
	\begin{tabular}{r|c||c|c|c|c|c||c|c|c|c||c|c}
		\multicolumn{2}{c||}{} & \multicolumn{5}{c||}{\bf \revz{Total bugs}} & \multicolumn{4}{c||}{\bf \revz{Data-race bugs}} \\
		% ---
		& {\bf Num.} & \multicolumn{2}{c|}{\bf \revz{\quicksand}} & \multicolumn{3}{c||}{\bf \revz{Control (SSS-MC)}}
		& \multicolumn{2}{c|}{\bf \revz{Limited HB}} & \multicolumn{2}{c||}{\bf \revz{Pure HB}} &
		{\bf Mutual} & {\bf Avg. tested} \\
		% ---
		{\bf Test} & {\bf tested} & {\bf LHB} & {\bf PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem}
		& {\bf All} & {\bf Nondet.} & {\bf \revz{All}} & {\bf \revz{Nondet.}}
		& {\bf timeouts} & {\bf subset SSes} \\
		% ---
		\hline
				% #test  qs-lhb  qs-phb           icb             dpor    everyw  dronly  nondets(LHB)    dronly(PHB)    nondets(LHB) mutualtimeout    comp.SSes
		{\tt bcast}	& 79	& 8	& 8		& 5		& 6	& 7	& 2	& \revz{1}	& \revz{2}	& \revz{1}	& 7		& 112.3	\\
		{\tt join} 	& 79	& 23	& 20		& 13		& 13 	& 14	& 11	& \revz{4}	& \revz{7}	& \revz{3}	& \revz{12}	& \revz{69.7}	\\
		{\tt mx} 	& 79	& 10	& 9		& 1		& 1	& 12	& 9	& \revz{1}	& \revz{8}	& \revz{1}	& 0		& -	\\
		{\tt sem} 	& 79	& 17	& 16		& 12		& 11	& 12	& 7	& \revz{3}	& \revz{6}	& \revz{2}	& \revz{50}	& \revz{77.4}	\\
		{\tt signal} 	& 79	& 10	& 8		& 5		& 5	& 11	& 6	& \revz{1}	& \revz{3}	& \revz{2}	& 45		& \revz{59.6}	\\
		{\tt rwlock} 	& 79	& 27	& 26		& 25		& 23	& 28	& 4	& \revz{1}	& \revz{3}	& \revz{0}	& \revz{44}	& \revz{86.3}	\\
		\hline
		{\tt sched} 	& 59	& 7	& 7		& 1		& 1	& 8	& 6	& \revz{4}	& \revz{6}	& \revz{6}	& 2		& 13.0	\\
		{\tt alarm} 	& 44	& 21	& 12		& 16		& 5	& 29	& 17	& \revz{1}	& \revz{7}	& \revz{6}	& \revz{17}	& \revz{7.8}	\\
		{\tt wait} 	& 52	& 30	& \revz{26}	& \revz{24}	& 23	& 1	& 7	& \revz{2}	& \revz{2}	& \revz{0}	& \revz{15}	& \revz{33.8}	\\
		\hline
		{\bf Total}	& 629	& 153	& \revz{132}	& \revz{102}	& 88	& 122	& 69	& \revz{15}	& \revz{44}	& \revz{21}	& \revz{192}	& \revz{65.8} 	\\
	\end{tabular}
		}
	\end{center}
	\caption{Summary of bugs
	%and data races
	found by each test program.
		\revisionx{QS-LHB and QS-PHB are \quicksand; ICB/DPOR/ShMem are the controls (\sect{\ref{sec:eval-setup}}).}
		``\revisionz{Data-race} bugs'' counts among \revisionz{\quicksand's} bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
	among those, \revisionz{the ``Nondeterministic'' columns show} how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
	``Mutual timeouts'' counts how often both \revisionx{QS-Limited-HB and SSS-MC-ICB} timed out with no bug found;
	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
		}
	\label{tab:drbugs}
\end{table*}

{\bf Finding new data-race bugs.}
\revision{Compared to \revisionx{SSS-MC-ICB (the fastest among the control experiments),} \quicksand~finds}~more bugs within any CPU budget greater than \revision{200}~seconds.
\revisionx{Compared to SSS-MC-Shared-Mem (the best SSS-MC approach in the long term), \quicksand's Limited HB version} ultimately concludes 10 CPU-hours with \revisionx{125\%} as many bugs in total.
\revision{Before the break-even point at 200 seconds,
\quicksand~lags behind \revisionx{SSS-MC-ICB} due to additional start-up overhead from its tenfold parallelism.
However, converting SSS-MC's early CPU-time advantage into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
Figure~\ref{fig:dowefindbugsfaster}(b) gives \quicksand~full credit for its inherent parallelism:
with a 10 core allocation, it outperforms SSS-MC
for any fixed budget of wall-clock time.}
\revisionx{After 1 hour of wall-clock time, tenfold \quicksand~performs 158\% as well as SSS-MC-ICB.}

@@

The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test program.
In \mxtest, in which we do not trust the lock implementation's correctness,
\revisionx{SSS-MC-ICB and SSS-MC-DPOR}
found dramatically fewer bugs (just 1)\footnote{
	The one bug SSS-MC found was in a fully-assembler lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient to find the bug.}.
Though it often suffices to assume correctly-implemented locks \cite{dbug-thesis},
we consider this strong evidence that new low-level synchronization code must be verified with data-race PPs.

% Future work (for thesis maybe?) Compare QS-ETAs to QS-Random to evaluate "smaller is better" claim.
{\bf Finding the same bugs faster.}
\revision{The QS-Sync-Only experiment tests whether}~Iterative Deepening is effective even for MC domains without data races.
\revision{When \quicksand~ignores all data-race candidates,
its results are competitive with SSS-MC-DPOR, but SSS-MC-ICB outperforms it.
This is unsurprising: the seed subsets of PPs
QS-Sync-Only is limited to
(\sect{\ref{sec:initial-pp}})
are much less flexible than ICB's preemption strategy (\sect{\ref{sec:overview-sssmc}}).
This result suggests that in future work, \quicksand~should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.}
%\footnote{
%Because \quicksand~is not yet instrumented to subset hard-coded PPs beyond the 4 ways shown in Figure~\ref{fig:id},
%we ran these tests for 2.5 hours on 4 CPUs each.
%Future work could parallelize QS-no-DR-PPs further; see \sect{\ref{sec:future}}.}.

@@

{\bf Partial verification.}
When a MC job times out, the user may prefer a brief summary of what parts of the test were verified, rather than writing off all the CPU time as a waste.
%Beyond using the state space estimator to guess the percent coverage,
%We know of no approach to quantify the probability
%that a bug would be exposed by an untested interleaving,
While recent work \cite{randomized-scheduler} attempts to quantify the probability
\revisionz{that a bug remains in some untested interleaving},
%that an untested interleaving would expose a bug,
\quicksand~instead reports which subsets of PPs resulted in state spaces that did complete in time.
%\quicksand~does not attempt any such quantitative guarantee,
%but can at least report
On \revisionz{229} tests, \revisionx{SSS-MC-ICB} timed out after 10 hours with no bugs found.
Among these tests, \quicksand~found bugs in \revisionz{37}.
For the other \revisionz{192}, we show the number of state spaces \quicksand~was able to complete in the ``Average tested subset SSes'' column of Table~\ref{tab:drbugs}.
These completions guarantee that, if the test program could expose a bug,
it would only be found by a new data-race PP not discovered yet, or by a superset combination of PPs not reached.
\revision{
%While obviously not as strong as full verification,
Prior work \cite{sealing} has argued the value of similar {\em compositional testing} when full verification is intractable,
deferring to the user's expertise to judge the value of each subset of PPs verified.
}

@@

%\quicksand's total verifications
%%before a given elapsed CPU consumption.
%\revision{against those provided by SSS-MC-Shared-Mem\footnote{
%	\revision{In kernel or thread library code, a thread's own stack accesses may not always be private. %assumption may not hold.
%	For example, many P2 condvars use stack-allocated list nodes to track waiting threads.
%	SSS-MC-Shared-Mem's verifications are only sound in the absence of such sharing!
%	\quicksand~detects these dynamically, without needing to discriminate between stack and heap.}
%	%Expanding the statically-coded PPs to include stack accesses would make SSS-MC-Shared-Mem's performance even worse.
%	% Especially for testing kernel-level code, deciding which of a thread's memory accesses are shared versus private is nonstraightforward, as threads may share data structures allocated on their own stacks for synchronization. Hence, the common heuristic of ignoring stack accesses as private (which we employ for this control experiment) introduces some unsoundndess.
%}.

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{totalverifs-v2.pdf}
	\caption{Cumulative distribution of tests fully verified
	\revisionx{by QS-Limited-HB, QS-Pure-HB, and SSS-MC-Shared-Mem}
	(\sect{\ref{sec:totalverif}}).
	%Some tests had no data-race candidates, and hence could also be verified by .}
	36 data-race-free tests were also soundly verified by QS-Sync-Only, SSS-MC-DPOR, and SSS-MC-ICB.}
	\label{fig:totalverif}
\end{figure}
